---
title: "Class 10: Halloween Mini-Project"
author: "Xiaoyan Wang(A16454055)"
format: pdf
editor: visual
---

## Exploratory Analysis of Halloween Candy

```{r}
candy_file<- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

-   **Q1**. How many different candy types are in this dataset?

    ```{r}
    nrow(candy)
    ```

-   **Q2**. How many fruity candy types are in the dataset?

    ```{r}
    sum(candy$fruity)
    ```

-   **Q3**. What is your favorite candy in the dataset and what is it’s `winpercent` value?

    ```{r}
    candy["Skittles original", ]$winpercent
    ```

-   **Q4**. What is the `winpercent` value for “Kit Kat”?

    ```{r}
    candy["Kit Kat", ]$winpercent
    ```

-   **Q5**. What is the `winpercent` value for “Tootsie Roll Snack Bars”?

    ```{r}
    candy["Tootsie Roll Snack Bars", ]$winpercent
    ```

```{r}
library(dplyr)

candy |>
  filter(rownames(candy)=="Haribo Happy Cola")|>
  select(winpercent)
```

Q: Find Fruity candy that have a winpercent \>= 50

```{r}
candy |>
  filter(winpercent> 50)|>
  filter(fruity==1)
```

```{r}
top.candy<- candy[candy$winpercent >50,]
top.candy[top.candy$fruity == 1,]
```

Quick overview of a given dataset:

```{r}
#install.packages("skimr")
library("skimr")
skim(candy)
```

Looks like the "winpercent" variable or column is masured on a different scale than everything else. I will need to scale my data before doing any analysis like PCA etc.

-   **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

    The winpercent is in 0-100 range, representing a percentage, whereas other columns from the dataset have a range of 0-1.

-   **Q7**. What do you think a zero and one represent for the `candy$chocolate` column?

    The 0 means this candy is not chocolate, and 1 means this candy is/contains chocolate

-   **Q8**. Plot a histogram of `winpercent` values

    We can do this in few ways. e.g. the "base" R 'hist()' function or with 'ggplot'.

    ```{r}
    hist(candy$winpercent, breaks = 100)
    ```

    ```{r}
    library(ggplot2)
    ggplot(candy)+
      aes(winpercent)+
      geom_histogram(binwidth =8)
    ```

-   **Q9**. Is the distribution of `winpercent` values symmetrical?

    No

-   **Q10**. Is the center of the distribution above or below 50%?

    ```{r}
    summary(candy$winpercent)
    ```

-   **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

    ```{r}
    candy |>
      filter(as.logical(fruity))
    ```

    ```{r}
    choc.candy <- candy %>% filter(as.logical(chocolate))
    fruit.candy <- candy %>% filter(as.logical(fruity))
    mean(choc.candy$winpercent, na.rm = TRUE)
    mean(fruit.candy$winpercent, na.rm = TRUE)
    ```

-   **Q12**. Is this difference statistically significant?

    ```{r}
    t.test(choc.candy$winpercent, fruit.candy$winpercent)
    ```

    Yes. The p-value is small, indicating that we can reject the null hypothesis and state there is significant differences between the mean between winpercent of fruity candy and chocolate candy.

-   **Q13**. What are the five least liked candy types in this set?

    ```{r}
    candy %>% arrange(winpercent) %>% head(5)
    ```

-   **Q14**. What are the top 5 all time favorite candy types out of this set?

    ```{r}
    candy %>% arrange(desc(winpercent)) %>% head(5)
    ```

-   **Q15**. Make a first barplot of candy ranking based on `winpercent` values.

    lets do a barplot

    ```{r}
    ggplot(candy) + 
      aes(winpercent, rownames(candy)) +
      geom_col()
    ```

-   **Q16**. This is quite ugly, use the `reorder()` function to get the bars sorted by `winpercent`?

    ```{r}
    ggplot(candy)+
      aes(x=winpercent,
          y= reorder(rownames(candy),winpercent),
          fill=chocolate)+
      geom_col()
    ```

    I want a more custom color scheme where I can see both chocolate and bar and fruity etc. all from the one plot. To do this, we can roll our own color vector...

    ```{r}
    mycol<- rep("black",nrow(candy))
    mycol[as.logical(candy$chocolate)] <- "chocolate"
    mycol[as.logical(candy$bar)] <- "red"
    mycol[as.logical(candy$fruity)] <- "pink"
    mycol[row.names(candy)=="Skittles original"] <- "blue"
    ```

    ```{r}
    ggplot(candy)+
      aes(x=winpercent,
          y= reorder(rownames(candy),winpercent),
          fill=chocolate)+
      geom_col(fill=mycol)
    ```

-   **Q17**. What is the worst ranked chocolate candy?

    Sixlets

-   **Q18**. What is the best ranked fruity candy?

    Starburst

plot of winpercent vs pricepercent to see what would be the candy to by

```{r}
mycol<- rep("black",nrow(candy))
mycol[as.logical(candy$chocolate)] <- "orange"
mycol[as.logical(candy$bar)] <- "blue"
mycol[as.logical(candy$fruity)] <- "red"
mycol[row.names(candy)=="Skittles original"] <- "skyblue"
```

```{r}
ggplot(candy)+
  aes(x= winpercent,
      y= pricepercent)+
  geom_point(col=mycol)
```

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycol) + 
  geom_text_repel(col=mycol, size=3.3, max.overlaps = 5)
```

-   **Q19**. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

    Tootsie Roll Midgies is the most bang for

    ```{r}
    ord <- order(-candy$pricepercent, decreasing = TRUE)
    head( candy[ord,c(11,12)], n=5 )
    ```

-   **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

    The least popular one is Nik L Nip

    ```{r}
    ord <- order(candy$pricepercent, decreasing = TRUE)
    head( candy[ord,c(11,12)], n=5 )
    ```

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

-   **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

    Fruity and Chocolate

-   **Q23**. Similarly, what two variables are most positively correlated?

    Chocolate and Winpercent/ Chocolate and bar

### Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[, 1], pca$x[, 2],
     xlab = "PC1", 
     ylab = "PC2",)
```

```{r}
plot(pca$x[,1:2], col=mycol, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycol)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=mycol, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#install.packages("plotly")
library(plotly)
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

-   **Q24**. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, pluribus and hard. These variables may be correlated to each other, like fruity candy usually being hard or pluribus.
